---
title: "Expenditure"
author: "Imali Punsara"
date: "`r Sys.Date()`"
output: html_document
---

# Load necessary libraries
```{r}
library(tidyverse)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(class)

```

# Load dataset
```{r}
library(readr)
data <- read_csv("healthexpenditurebyareaandsource.csv")
View(healthexpenditurebyareaandsource)
```

# Exploratory Data Analysis (EDA)
```{r}
print(str(data))  # Check structure of dataset
print(summary(data))  # Summary statistics
print(colSums(is.na(data)))  # Check for missing values

```

# Convert categorical variables to factors
```{r}
data$state <- as.factor(data$state)
data$area_of_expenditure <- as.factor(data$area_of_expenditure)
data$broad_source_of_funding <- as.factor(data$broad_source_of_funding)
data$detailed_source_of_funding <- as.factor(data$detailed_source_of_funding)
```

# Define target variable: Classify expenditure source as Government vs Non-Government
```{r}
data$Funding_Category <- ifelse(data$broad_source_of_funding == "Government", "Gov", "Non-Gov")
data$Funding_Category <- as.factor(data$Funding_Category)

```

# Remove non-numeric columns not used for classification
```{r}
numeric_cols <- c("real_expenditure_millions")
data_clean <- data %>% select(numeric_cols, Funding_Category)

```

# Split data into training and testing sets
```{r}
set.seed(123)
trainIndex <- createDataPartition(data_clean$Funding_Category, p=0.7, list=FALSE)
trainData <- data_clean[trainIndex, ]
testData <- data_clean[-trainIndex, ]

```

# Train Classification Model - Decision Tree
```{r}
set.seed(123)
model_dt <- rpart(Funding_Category ~ ., data=trainData, method="class")
```

# Visualize Decision Tree
```{r}
rpart.plot(model_dt)
```

# Predictions - Decision Tree
```{r}
predictions_dt <- predict(model_dt, testData, type="class")

```

# Evaluate Decision Tree Model Performance
```{r}
confusionMatrix(predictions_dt, testData$Funding_Category)
```

# Train Classification Model - Random Forest
```{r}
set.seed(123)
model_rf <- randomForest(Funding_Category ~ ., data=trainData, ntree=100)

```

# Predictions - Random Forest
```{r}
predictions_rf <- predict(model_rf, testData)
```

# Evaluate Random Forest Model Performance
```{r}
confusionMatrix(predictions_rf, testData$Funding_Category)
```

# Train Classification Model - KNN
```{r}
set.seed(123)
k <- 5  # Number of neighbors
train_scaled <- scale(trainData[, numeric_cols])
test_scaled <- scale(testData[, numeric_cols])
predictions_knn <- knn(train_scaled, test_scaled, cl=trainData$Funding_Category, k=k)

```

# Evaluate KNN Model Performance
```{r}
confusionMatrix(predictions_knn, testData$Funding_Category)

```

# Clustering - K-Means
```{r}
set.seed(123)
 data_scaled <- scale(data_clean[, numeric_cols])
```

# Determine optimal number of clusters using the Elbow method
```{r}
fviz_nbclust(data_scaled, kmeans, method = "wss")
```

# Apply K-Means with 3 clusters
```{r}
kmeans_result <- kmeans(data_scaled, centers=3, nstart=25)
data_clean$Cluster <- as.factor(kmeans_result$cluster)
```

# Visualizing Clusters
```{r}
fviz_cluster(list(data = data_scaled, cluster = kmeans_result$cluster), geom = "point", ellipse.type = "convex")
}else {
  print("Error: Not enough numeric variables for clustering visualization.")
}

```

#Clustering - Hierarchical Clustering
```{r}
if (ncol(data_clean[, numeric_cols, drop=FALSE]) > 1) {
  hc <- hclust(dist(data_scaled), method = "ward.D2")
  plot(hc, main = "Dendrogram of Hierarchical Clustering", sub = "", xlab = "")
  rect.hclust(hc, k=3, border="red")
} else {
  print("Error: Not enough numeric variables for hierarchical clustering.")
}
```

#DBSCAN (Density-Based Clustering)
```{r}
library(dbscan)

# Standardize numeric data
data_scaled <- scale(data_clean[, numeric_cols])

# Apply DBSCAN clustering
set.seed(123)
dbscan_result <- dbscan(data_scaled, eps = 0.5, minPts = 5)

# Add cluster labels to data set
data_clean$Cluster_DBSCAN <- as.factor(dbscan_result$cluster)

# Visualize Clustering
ggplot(data_clean, aes(x = real_expenditure_millions, y = Cluster_DBSCAN, color = Cluster_DBSCAN)) +
  geom_jitter() + 
  labs(title = "DBSCAN Clustering of Health Expenditure", x = "Expenditure (Millions)", y = "Cluster")

```

#Gaussian Mixture Models (GMM)
```{r}
library(mclust)

# Standardize numeric data
data_scaled <- scale(data_clean[, numeric_cols])

# Fit GMM model
set.seed(123)
gmm_model <- Mclust(data_scaled)

# Add cluster labels
data_clean$Cluster_GMM <- as.factor(gmm_model$classification)

# Visualize Clustering
fviz_cluster(list(data = data_scaled, cluster = gmm_model$classification), geom = "point", ellipse.type = "convex")

```

# Clustering - Gaussian Mixture Model (GMM)
```{r}
set.seed(123)
gmm_model <- Mclust(data_scaled)
data_clean$Cluster_GMM <- as.factor(gmm_model$classification)
```

# Visualizing GMM Clustering
```{r}
fviz_cluster(list(data = data_scaled, cluster = gmm_model$classification), geom = "point", ellipse.type = "convex")
```


#Mean Shift Clustering
```{r}
# Load necessary libraries
library(ggplot2)
library(mlbench)

# Sample data set
data(data_scaled)

# Mean Shift clustering
mean_shift_result <- meanShift(data[, -5], bandwidth = 0.5)

# Add cluster results to the original dataset
data$Cluster <- as.factor(mean_shift_result$cluster)

# Plot the clusters
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Cluster)) +
  geom_point() +
  labs(title = "Mean Shift Clustering on Iris Dataset")

```

