---
title: "Data Mining Assignment"
author: "Imali Punsara"
date: "`r Sys.Date()`"
output: html_document
---

# Load necessary libraries
```{r}
library(tidyverse)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(class)
library(cluster)
library(factoextra)
```

#Load dataset
```{r}
summaryg21 <- read_csv("C:/Users/singhagiri/Desktop/DM/summaryg21.csv")
View(summaryg21)
```

#Rename columns
```{r}
colnames(summaryg21) <- c("Sector", "Output_M", "Employment_Jobs", "Wages_Salaries_M", "Local_Sales_M",
                    "Regional_Exports_M", "Local_Expenditure_M", "Regional_Imports_M", "Value_Added_M")
```

#Convert numeric columns to proper format
```{r}
data <- summaryg21 %>% mutate(across(-Sector, ~as.numeric(gsub("[$,]", "", .))))
```

```{r}
numeric_cols <- c("Output_M", "Employment_Jobs", "Wages_Salaries_M", "Local_Sales_M", 
                  "Regional_Exports_M", "Local_Expenditure_M", "Regional_Imports_M", "Value_Added_M")

```

#check for missing values
```{r}
missing_values <- colSums(is.na(summaryg21))
print(missing_values)
```

```{r}
if (is.data.frame(summaryg21)) {
  missing_values <- colSums(is.na(summaryg21))
  print(missing_values)
} else {
  print("Error: Data is not in a valid format for checking missing values.")
}
```

#summary statistics
```{r}
summary(summaryg21)

```

```{r}
names(summaryg21)
head(summaryg21)
tail(summaryg21)
str(summaryg21)
```


#check the dimension and number of points 
```{r}
nrow(summaryg21)
ncol(summaryg21)
dim(summaryg21) 
```


#Visualize missing data
```{r}
if (length(missing_values) > 0) {
  ggplot(summaryg21 = as.data.frame(missing_values), aes(x = names(missing_values), y = missing_values)) + 
    geom_bar(stat = "identity", fill = "red") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    labs(title = "Missing Values in Dataset", x = "Columns", y = "Count")
}
```

```{r}
missing_df <- data.frame(Variable = names(missing_values), Missing_Count = missing_values)
ggplot(missing_df, aes(x = Variable, y = Missing_Count)) + 
  geom_bar(stat = "identity", fill = "red") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  labs(title = "Missing Values in Dataset", x = "Columns", y = "Count")
```

#data visualization
## Histogram for distribution of numerical features
```{r}
ggplot(summaryg21, aes(x = Output_M)) + geom_histogram(binwidth = 50, fill = "blue", alpha = 0.7) +
  labs(title = "Distribution of Output", x = "Output ($M)", y = "Count")
```

```{r}
ggplot(summaryg21, aes(x = as.numeric(Output_M))) +
  geom_histogram(binwidth = 50, fill = "blue", alpha = 0.7) +
  labs(title = "Distribution of Output", x = "Output ($M)", y = "Count")

```

```{r}
ggplot(summaryg21, aes(x = as.numeric(as.character(Wages_Salaries_M)))) +
  geom_histogram(binwidth = 50, fill = "blue", alpha = 10) +
  labs(title = "Distribution of wages_Salaries_M", x = "Wages_Salaries_M", y = "Count")

```


```{r}
hist(summaryg21)
```

## Boxplot for detecting outliers
```{r}
ggplot(summaryg21, aes(y = Output_M)) + geom_boxplot(fill = "orange") +
  labs(title = "Boxplot of Output", y = "Output ($M)")
```

## Scatter plot to examine correlations
```{r}
ggplot(summaryg21, aes(x = Output_M, y = Employment_Jobs)) + geom_point(color = "red") +
  labs(title = "Scatter Plot: Output vs Employment", x = "Output ($M)", y = "Employment (Jobs)")
```

## Correlation heatmap
```{r}
cor_matrix <- cor(summaryg21[numeric_cols], use = "complete.obs")
ggcorrplot(cor_matrix, lab = TRUE)

```

```{r}
numeric_cols <- sapply(summaryg21, is.numeric)
numeric_data <- summaryg21[, numeric_cols]
correlation_matrix <- cor(numeric_data, use = "complete.obs")
```

```{r}
ggplot(summaryg21, aes(x = Employment_Jobs)) +
  geom_bar(fill = "purple") +
  labs(title = "Count of Employment Levels", x = "Employment_jobs", y = "Count") +
  theme_minimal()
```

```{r}
ggplot(summaryg21, aes(x = Output_M)) +
  geom_histogram(binwidth = 50, fill = "blue", alpha = 0.7) +
  labs(title = "Distribution of Output", x = "Output ($M)", y = "Count") +
  theme_minimal()

```

```{r}
str(summaryg21)

```

```{r}
# Convert Output_M to numeric, ensuring no characters remain
summaryg21$Output_M <- as.numeric(gsub("[$,]", "", summaryg21$Output_M))

# Check for NA values after conversion
sum(is.na(summaryg21$Output_M))

```

```{r}
# Histogram for distribution of Output_M
ggplot(summaryg21, aes(x = Output_M)) +
  geom_histogram(binwidth = 50, fill = "blue", alpha = 0.7) +
  labs(title = "Distribution of Output", x = "Output ($M)", y = "Count") +
  theme_minimal()

```

```{r}
cor_matrix <- cor(summaryg21[numeric_cols], use = "complete.obs")
ggcorrplot(cor_matrix, lab = TRUE) +
  labs(title = "Correlation Heatmap") +
  theme_minimal()
```

```{r}
library(ggcorrplot)

```

```{r}
cor_matrix <- cor(summaryg21[numeric_cols], use = "complete.obs")
ggcorrplot(cor_matrix, lab = TRUE) +
  labs(title = "Correlation Heatmap") +
  theme_minimal()
```

# Visualize target variable distribution (replace 'target_column' with actual column name)
```{r}
ggplot(summaryg21, aes(x = Employment_Jobs)) +
  geom_bar() +
  theme_minimal() +
  ggtitle("Target Variable Distribution")
```
# Splitting data into training and testing sets
```{r}
set.seed(123)
trainIndex <- createDataPartition(summaryg21$Employment_Jobs, p = 0.8, list = FALSE)
trainData <- summaryg21[trainIndex, ]
testData <- summaryg21[-trainIndex, ]
```

# Define target variable (replace with a suitable column name based on dataset context)
```{r}
target_variable <- "Employment_Jobs" 
```

```{r}
summaryg21[[target_variable]] <- as.factor(summaryg21[[target_variable]])
```

# Visualize target variable distribution
```{r}
ggplot(summaryg21, aes(x = summaryg21[[target_variable]])) +
  geom_bar() +
  theme_minimal() +
  ggtitle("Target Variable Distribution")
```

# Splitting data into training and testing sets
```{r}
set.seed(123)
trainIndex <- createDataPartition(summaryg21[[target_variable]], p = 0.8, list = FALSE)
trainData <- summaryg21[trainIndex, ]
testData <- summaryg21[-trainIndex, ]

```

```{r}
library(dplyr)
 set.seed(2345)
 train.summary = sample_frac(summaryg21, 0.7)
 test.summary = summaryg21[-as.numeric(rownames(train.summary)),] # use as.numeric because rownames() returns character
```



```{r}
summaryg21$Employment_Jobs <- factor(ifelse(summaryg21$Employment_Jobs >= 1000, "High", "Low"))
```

```{r}
set.seed(123)
model_dt <- rpart(Employment_Level ~ Output_M + Wages_Salaries_M + Local_Sales_M + Regional_Exports_M,
                  summaryg21 = trainData, method="class")
```

```{r}
confusionMatrix(as.factor(predictions_dt), testData$Employment_Level)
```











